---
title: "In-Hospital Mortality Prediction"
author: "Katya Pandey 19BCE1312"
date: '2022-04-24'
output: html_document
---

## 1. Loading the dataset

```{r}
options(max.print=1000000)
# Reading the dataset
df <- read.csv("data01.csv", fileEncoding = "UTF-8", na.strings = "..")
head(df)
```

```{r}
# Parameters affecting patient mortality
print("Column names present in dataset: ")
names(df)
```

```{r}
# The structure of the data 
str(df)
```

```{r}
# Summary of the dataset
summary(df)
```

## 2. Data Pre-processing

#### Data Cleaning
```{r}
# Check for null or missing values in the dataset
sum(is.na(df))
# Check for null or missing values in each column of the dataset
print("Column-wise presence of missing data: ")
colSums(is.na(df))
```

###### Inference: We have a total of 1929 null values in the dataset.

```{r}
# Plotting the number of missings for each variable using gg_miss_var() funtion
#Importing the necessary libraries
library(naniar)
gg_miss_var(df)
```

```{r}
#PCO2 and PH have many missing values. Analyzing the columns further by creating a linear regression model using lm() function, we’ll get the summary output using the summary() function. 
summary(lm(PCO2~.,data=df))
summary(lm(PH~.,data=df))
```

###### We are not omitting any rows and columns as it we do not want any loss of data.

```{r}
# Linear Regression Imputation
# Importing the necessary libraries
library(simputation)
df$PCO2<-as.numeric(df$PCO2)
imp_df <- impute_lm(df[,-c(1,2)],PCO2~as.numeric(outcome)+Platelets+PH)
imp_df <- impute_lm(imp_df,PH~gendera+temperature+Creatinine+Bicarbonate+PCO2)
```

###### Some of the values are not taken into account. The linear model does not give an output as one of the predictor variables is missing. 

```{r}
#Importing the necessary libraries
library(imputeTS)
imp_df$outcome<-as.factor(imp_df$outcome)
imp_df<-na_mean(imp_df[,-c(45,48)])
```

```{r}
imp_df$PH<-df$PH
imp_df$PCO2<-df$PCO2
```

```{r}
imp_df <- impute_lm(df[,-c(1,2)],PCO2~as.numeric(outcome)+Platelets+PH)
imp_df <- impute_lm(imp_df,PH~gendera+temperature+Creatinine+Bicarbonate+PCO2)
```

```{r}
imp_df<-na.omit(imp_df)
# Check for null or missing values in the dataset after imputation
sum(is.na(imp_df))
# Plotting the missing dataset values if any
gg_miss_var(imp_df)
```

###### As we can see, there are no null values after the data imputation.

```{r}
normalize <- function(x) {
    return((x - min(x)) / (max(x) - min(x)))
}
imp_df$outcome<-as.numeric(imp_df$outcome)
imp_df<-apply(imp_df,2,normalize)
imp_df<-as.data.frame(imp_df)
```

```{r}
# Changing data values for better understanding
#imp_df$outcome[imp_df$outcome == 0] <- "alive"
#imp_df$outcome[imp_df$outcome == 1] <- "dead"
head(imp_df)
str(imp_df)
```

## 3. Exploratory Data Analysis

```{r}
#Importing the necessary libraries
library("ggplot2")
library("ggpubr")
theme_set(theme_pubr())
```

```{r}
# Importing the necessary packages
library(funModeling) 
library(tidyverse) 
```

```{r}
# Correlation between mortality outcome and all other parameters
cor(imp_df$outcome,imp_df$age)
cor(imp_df$outcome,imp_df$gendera)
cor(imp_df$outcome,imp_df$BMI)
cor(imp_df$outcome,imp_df$hypertensive)
cor(imp_df$outcome,imp_df$atrialfibrillation)
cor(imp_df$outcome,imp_df$CHD.with.no.MI)
cor(imp_df$outcome,imp_df$diabetes)
cor(imp_df$outcome,imp_df$deficiencyanemias)
cor(imp_df$outcome,imp_df$depression)
cor(imp_df$outcome,imp_df$Hyperlipemia)
cor(imp_df$outcome,imp_df$Rel.failure)
cor(imp_df$outcome,imp_df$heart.rate)
cor(imp_df$outcome,imp_df$Systolic.blood.pressure)
cor(imp_df$outcome,imp_df$Diastolic.blood.pressure)
cor(imp_df$outcome,imp_df$Respiratory.rate)
cor(imp_df$outcome,imp_df$temperature)
cor(imp_df$outcome,imp_df$SP.O2)
cor(imp_df$outcome,imp_df$Urine.output)
cor(imp_df$outcome,imp_df$hematocrit)
cor(imp_df$outcome,imp_df$RBC)
cor(imp_df$outcome,imp_df$MCH)
cor(imp_df$outcome,imp_df$MCHC)
cor(imp_df$outcome,imp_df$MCV)
cor(imp_df$outcome,imp_df$RDW)
cor(imp_df$outcome,imp_df$Leucocyte)
cor(imp_df$outcome,imp_df$Platelets)
cor(imp_df$outcome,imp_df$Neutrophils)
cor(imp_df$outcome,imp_df$Basophils)
cor(imp_df$outcome,imp_df$Lymphocyte)
cor(imp_df$outcome,imp_df$PT)
cor(imp_df$outcome,imp_df$INR)
cor(imp_df$outcome,imp_df$NT.proBNP)
cor(imp_df$outcome,imp_df$Creatine.kise)
cor(imp_df$outcome,imp_df$Creatinine)
cor(imp_df$outcome,imp_df$Urea.nitrogen)
cor(imp_df$outcome,imp_df$glucose)
cor(imp_df$outcome,imp_df$Blood.potassium)
cor(imp_df$outcome,imp_df$Blood.sodium)
cor(imp_df$outcome,imp_df$Blood.calcium)
cor(imp_df$outcome,imp_df$Chloride)
cor(imp_df$outcome,imp_df$BMI)
cor(imp_df$outcome,imp_df$Anion.gap)
cor(imp_df$outcome,imp_df$Magnesium.ion)
cor(imp_df$outcome,imp_df$PH)
cor(imp_df$outcome,imp_df$Bicarbote)
cor(imp_df$outcome,imp_df$Lactic.acid)
cor(imp_df$outcome,imp_df$PCO2)
cor(imp_df$outcome,imp_df$EF)
```
#### Maximum correlation is between outcome and BMI.

#### Visualization of patient attributes

###### Count of patients who are alive and dead 
```{r}
t <- table(df$outcome)
t <- as.data.frame(t)
colnames(t) <- c("aliveness","count")
ggplot(t, aes(x=aliveness, y=count, fill=aliveness)) +
geom_bar(stat="identity", color="black") +
theme_light() +
geom_text(aes(label=count), vjust=-0.4, size=4) +
scale_fill_brewer(palette="Set2")
```
###### 159 patients died while admitted in the ICU.

#### Visualization of frequencies of all parameters affecting mortality


```{r}
g4 <- table(df$atrialfibrillation)
g4 <- as.data.frame(g4)
colnames(g4) <- c("atrialfibrillation","count")

ggplot(g4, aes(x=atrialfibrillation, y=count, fill=atrialfibrillation)) +
geom_bar(stat="identity", color="black") +
theme_minimal() +
geom_text(aes(label=count), vjust=-0.4, size=4) +
scale_fill_brewer(palette="Set3")
```

```{r}
x <- table(df$gendera)
labels <-  c("male","female")
piepercent<- round(100*x/sum(x), 1)
pie(x, labels = piepercent, main = "Age-based distribution",col = rainbow(length(x)))
legend("topright", c("male","female"), cex = 0.8,
   fill = rainbow(length(x)))
```
#### More number of female patients than male patients.

```{r}
g3 <- table(df$hypertensive)
g3 <- as.data.frame(g3)
colnames(g3) <- c("hypertensive","count")

ggplot(g3, aes(x=hypertensive, y=count, fill=hypertensive)) +
geom_bar(stat="identity", color="black") +
theme_minimal() +
geom_text(aes(label=count), vjust=-0.4, size=4) +
scale_fill_brewer(palette="Set1")
```

```{r}
hist(df$Anion.gap, xlab = "Anion-Gap", ylab = "Count", col = "pink",border = "blue", main="Frequency of Anion Gap in patients")
```

```{r}
ggplot(df, aes(x = age, y = RBC)) +
    geom_point(aes(color = "RBC"))
```


```{r}
library(plotrix)
x <- table(df$diabetes)
labels <-  c("no","yes")
piepercent<- round(100*x/sum(x), 1)
pie3D(x,labels = labels, explode = 0.1, main = "Count of patients with Diabetes",col = rainbow(length(x)))
legend("topright", c("no","yes"), cex = 0.8,
   fill = rainbow(length(x)))
```
```{r}
ggplot(df, aes(x = age, y = heart.rate)) +
    geom_point(aes(color = "heart.rate"))
```


```{r}
ggplot(df, aes(x = age, y = temperature)) +
    geom_point(aes(color = "temperature"))
```
```{r}
hist(df$BMI, xlab = "BMI",col = "yellow",border = "blue", main="Frequency of BMI values in patients")
```
```{r}
#### Depression-based patient count 
g1 <- table(df$depression)
g1 <- as.data.frame(g1)
colnames(g1) <- c("depression","count")
ggplot(g1, aes(x=depression, y=count, fill=depression)) +
geom_bar(stat="identity", color="black") +
theme_minimal() +
geom_text(aes(label=count), vjust=-0.4, size=4) +
scale_fill_brewer(palette="Set1")
```
```{r}
x <- table(df$Rel.failure)
labels <-  c("no","yes")
piepercent<- round(100*x/sum(x), 1)
pie3D(x,labels = labels, explode = 0.1, main = "Count of patients with Renal Failure",col = rainbow(length(x)))
legend("topright", c("no","yes"), cex = 0.8,
   fill = rainbow(length(x)))
```


## 4. Implementation of the prediction models

```{r}
# Importing the necessary libraries
library(caret)
library(e1071)
library(caTools)
library(kernlab)
library(stats19)
library(dplyr)
library(randomForest)
```

```{r}
# Splitting the data into training and testing data
# Split ratio is taken as 0.75
sample <- sample.split(imp_df,SplitRatio = 0.75)
training_data <- subset(imp_df,sample==TRUE)
testing_data <- subset(imp_df,sample==FALSE)
```

#### Model 1: Naive Bayes Model

```{r}
set.seed(234)
# Training the model
model <- naiveBayes(outcome~., data=training_data)
model
```

```{r}
# Predicting on test data
pred <- predict(model, testing_data)
pred
```

```{r}
# Confusion matrix
cm <- table(testing_data$outcome, pred)
cm
```

```{r}
# Model Evaluation
confusionMatrix(cm)
```

## Naive Bayes Accuracy: 91.38%

## Model 2: Logistic Regression
```{r}
# Training the model
logistic_model <- glm(outcome~., 
                      data = training_data, 
                      family = "binomial")
logistic_model
```

```{r}
plot(logistic_model)
```


```{r}
# Summary of the model
summary(logistic_model)
```

```{r}
# Predicting on test data
predict_reg <- predict(logistic_model, 
                       testing_data, type = "response")
predict_reg
```

```{r}
# Changing probabilities
predict_reg <- ifelse(predict_reg >0.5, 1, 0)
predict_reg
```

```{r}
# Evaluating model accuracy using confusion matrix
cm <- table(testing_data$outcome, predict_reg)
cm
```

```{r}
# Confusion matrix 
confusionMatrix(cm)
```

## Logistic Regression accuracy: 87.07%

## Model 3: Support Vector Machine
```{r}
set.seed(23) # for reproducibility
svm <- train(as.factor(outcome) ~ .,
data = training_data,
method = "svmRadial",
trControl = trainControl(method = "cv", number = 5),
tuneLength = 8
)

svm
```

```{r}
ggplot(svm) + theme_dark()
```

```{r}
ctrl <- trainControl(
method = "cv",
number = 15,
classProbs = TRUE,
summaryFunction = twoClassSummary # also needed for AUC/ROC
)
ctrl
```

```{r}
training_data$outcome<-as.factor(training_data$outcome)
levels(training_data$outcome) <- c("alive", "dead") 
```

```{r}
# Tune an SVM
set.seed(23) # for reproducibility
outcome_svm_auc <- train(
outcome ~ .,
data = training_data,
method = "svmRadial",
metric = "ROC", # area under ROC curve (AUC)
trControl = ctrl,
tuneLength = 15)

outcome_svm_auc
```

```{r}
confusionMatrix(outcome_svm_auc)
```
```{r}
# Arranging variables in terms of importance
prob_alive <- function(object, newdata) {
predict(object, newdata = newdata, type = "prob")[, "dead"]
}
```


```{r}
# Importing necessary libraries
library(vip)
set.seed(2827) # for reproducibility
# Visualizing variables in terms of their importance
vip(outcome_svm_auc, method = "permute", nsim = 5, train = training_data,
target = "outcome", metric = "auc", reference_class = "dead",
pred_wrapper = prob_alive)
```

```{r}
pred <- predict(outcome_svm_auc, testing_data)
pred
```

```{r}
testing_data$outcome<-as.factor(testing_data$outcome)
levels(testing_data$outcome) <- c("alive", "dead")
confusionMatrix(table(pred,as.factor(testing_data$outcome)))
```
## Support Vector Machine Accuracy: 88.79%

## Model-4: Linear Regression
```{r}
#imp_df$outcome<-as.numeric(imp_df$outcome)
mdl <- lm(outcome~., data = training_data)
mdl$coefficients
summary(mdl)
```

```{r}
plot(mdl)
```

```{r}
pred <- predict(mdl, testing_data)
# Changing probabilities
pred <- ifelse(pred > 0.5, 1, 0)
pred
```

```{r}
# Evaluating model accuracy using confusion matrix
cm <- table(testing_data$outcome, pred)
cm
```

```{r}
confusionMatrix(cm)
```
## Linear Regression Accuracy: 87.93%

## Model 5: Random Forest
```{r}
# imp_df$outcome[imp_df$outcome == 0] <- "alive"
# imp_df$outcome[imp_df$outcome == 1] <- "dead"
# Training the model
rfm <- randomForest(outcome~.,data=training_data, importance=T, proximity=T)
rfm
```

```{r}
plot(rfm)
```


```{r}
pred <- predict(rfm,testing_data)
pred <- ifelse(pred >0.5, 1, 0)
pred
```

```{r}
# Evaluating model accuracy using confusion matrix
cm <- table(testing_data$outcome, pred)
cm
```
```{r}
confusionMatrix(cm)
```
## Random Forest Accuracy: 92.24% 

## Comparative analysis of all the prediction models
```{r}
anova(rfm, mdl)
```

